---
title: "16S_Metagenomic_data_analysis"
output: html_notebook
---

# 16S Metagenomic Data Analysis - Overview

This R Markdown document performs comprehensive 16S rRNA gene amplicon sequencing analysis using the DADA2 pipeline for three distinct experimental datasets: **enrichment testing** (n=20), **host-DNA depletion testing** (n=60), and **DNA extraction kit comparison** (n=14), totaling **96 samples**. The workflow processes raw FASTQ sequences through quality filtering, denoising, chimera removal, and taxonomic classification using the SILVA reference database (v138.2), ultimately generating a phyloseq object containing OTU tables, sample metadata, and taxonomic assignments from Kingdom through Species level.

## Complete Dataset (All 96 Samples)

-   **Total samples**: 96 samples across all experimental conditions
-   **Sequencing**: Paired-end Illumina 16S rRNA gene amplicon sequencing (V3-V4 region)
-   **Primary phyloseq object**: `ps` - contains all 96 samples with complete OTU abundance table and taxonomy
-   **Quality control**: Read tracking through DADA2 pipeline stored in `track` dataframe (input → filtered → denoised → merged → non-chimeric reads)
-   **Taxonomic database**: SILVA nr99 v138.2 training set for species-level assignment
-   **Key objects**:
    -   `seqtab. nochim`: ASV (Amplicon Sequence Variant) table with chimeras removed
    -   `taxa`: Taxonomic classification table (Kingdom → Phylum → Class → Order → Family → Genus → Species)
    -   `ps_species_level_all96`: All samples agglomerated at Species level
    -   `ps_top20_rel_abund_all96`: Top 20 most abundant species across all 96 samples

------------------------------------------------------------------------

## DADA2 Pipeline Parameters

-   **Quality filtering**: `truncLen = c(230, 200)` for forward and reverse reads
-   **Error parameters**: `maxN = 0`, `maxEE = c(2,2)`, `truncQ = 2`
-   **Chimera removal**: `removeBimeraDenovo()` with default consensus method
-   **Taxonomic assignment**: `assignTaxonomy()` using SILVA nr99 v138.2 to species level
-   **Sequence tracking**: Pipeline step-wise read counts stored in `track` dataframe for QC

------------------------------------------------------------------------

## Dataset 1: Enrichment Testing Analysis (20 samples)

**Objective**: Evaluate microbial community dynamics during selective enrichment for foodborne pathogens.

### Sample Characteristics

-   **Sample size**: 20 samples (subset of 96-sample dataset)
-   **Enrichment broths**: Four types tested
    -   **BF Bolton base** (BFBB): Buffered peptone water with Bolton broth base
    -   **BF Bolton base w CAT**: Bolton base with Cefoperazone-Amphotericin-Tetrathionate supplement (antibiotic selection for *Campylobacter*)
    -   **BPW**: Buffered Peptone Water (non-selective pre-enrichment)
    -   **RVS**: Rappaport-Vassiliadis Soya broth (Salmonella-selective with malachite green)
-   **Time course**: 0, 4, 8, 12, and 24 hours of incubation at 37°C (5 timepoints × 4 broths = 20 samples)
-   **Sample codes**: BFBB0-24, CAT0-24, BPW0-24, RVS0-24

### Key Phyloseq Objects

-   `ps_enrichment_testing`: Complete enrichment dataset (20 samples) with metadata (Broth, Incubation, Incubation_hr)
-   `ps_family_level_enrichment`: Enrichment samples agglomerated at Family level
-   `ps_top20_families_enrichment`: Top 20 most abundant families in enrichment testing
-   `ps_rvs_only`: RVS broth subset only (5 samples, timepoints 0-24h)
-   `ps_rvs_bpw`: RVS and BPW broths only (10 samples), excludes Bolton-based media
-   `ps_genus_rvs`: RVS samples agglomerated at Genus level
-   `ps_all_broths`: All enrichment samples with complete, cleaned metadata (pathogen-focused analyses)

### Target Pathogens

-   **Campylobacterota** (Phylum): Contains *Campylobacter jejuni* and *C. coli*, leading bacterial cause of gastroenteritis worldwide
-   **Enterobacteriaceae** (Family): Contains *Salmonella* spp. and pathogenic *E. coli*, major foodborne pathogens regulated by food safety agencies

### Metadata Variables

-   `Broth`: Enrichment medium type (factor with 4 levels)
-   `Incubation`: Time with units (character: "0 hours", "4 hours", etc.)
-   `Incubation_hr`: Numeric incubation time (0, 4, 8, 12, 24) for plotting
-   `Sample`: Original sample code (BFBB0, CAT4, etc.)

------------------------------------------------------------------------

## Dataset 2: Host-DNA Depletion Testing (60 samples)

**Objective**: Evaluate methods for reducing chicken host DNA contamination in metagenomic samples.

### Sample Characteristics

-   **Sample size**: 60 samples
-   **Host-DNA depletion**: Two treatments
    -   **Host-DNA depletion** (H): Saponin-based host cell lysis + DNase treatment
    -   **No Host-DNA depletion** (NH): Standard processing without depletion
-   **Sample processing methods**: Three filtration approaches
    -   **Cheesecloth** (Ch): Coarse filtration through sterile cheesecloth
    -   **Filtered** (F): 0.45 µm syringe filtration
    -   **Filtered Centrifuge** (FCe): Filtration followed by centrifugation (10,000 × g)
-   **Enrichment conditions**: Three broth types tested
    -   **BFBB w CAT**: BF Bolton base with CAT supplement (*Campylobacter*-selective)
    -   **RVS**: Rappaport-Vassiliadis Soya (*Salmonella*-selective)
    -   **No Incubation**: Direct processing without enrichment
-   **Dilution**: Diluted vs. Not Diluted (ND suffix indicates no dilution)
-   **Sample types**: Biological replicates (1, 2, 3) and procedural blanks (B/Blank)
-   **Sample codes**: HCh1-3, HChB, HF1-3, HFB, HFCe1-3, HFCeB, NHCh1-3, etc. with ND suffix for undiluted

### Key Phyloseq Objects

-   `ps_hostDNAdep`: Complete host-DNA depletion dataset (60 samples) with metadata
-   `ps_hostDNAdep. top20_family`: Top 20 families agglomerated at Family level
-   `ps_no_cheesecloth`: Filtered and Filtered Centrifuge samples only (excludes Cheesecloth)
-   `ps_cheesecloth`: Cheesecloth samples only (excludes other filtration methods)
-   `ps_cheesecloth_incubation`: Cheesecloth samples that underwent enrichment (excludes "No Incubation")

### Metadata Variables

-   `Host-DNA Depletion`: Treatment status (Host-DNA depletion vs. No Host-DNA depletion)
-   `Method of Filtering`: Filtration approach (Cheesecloth, Filtered, Filtered Centrifuge)
-   `Incubation`: Enrichment broth type (BFBB w CAT, RVS, No Incubation)
-   `Dilution`: Dilution status (Diluted vs. Not Diluted)
-   `Sample Number/Type`: Replicate number (1, 2, 3) or Blank
-   `Sample_both`: Combined label (Sample Number + Dilution) for plotting

------------------------------------------------------------------------

## Dataset 3: DNA Extraction Kit Comparison (14 samples)

**Objective**: Compare DNA yield and microbial community recovery across three commercial DNA extraction kits.

### Sample Characteristics

-   **Sample size**: 14 samples
-   **DNA extraction kits**: Three commercial kits tested
    -   **Presto Mini gDNA**: Presto Mini gDNA Bacteria Kit (Geneaid)
    -   **Qiagen DNeasy Blood & Tissue**: DNeasy Blood & Tissue Kit (Qiagen)
    -   **ZymoResearch DNA Miniprep**: Quick-DNA Miniprep Kit (Zymo Research)
-   **Sample types**: Three categories
    -   **Chicken Rinse**: Chicken carcass rinse water (biological samples, n=3 per kit = 9 samples)
    -   **Spiked**: Positive controls spiked with known bacterial culture (n=1 per kit = 3 samples)
    -   **Unspiked**: Negative extraction controls (n=1 per kit = 2 samples, note: Presto has unspiked only)
-   **Sample codes**: PrMigDNA1-3, PRMgDNAUn; QDnBT1-3, QDnBTSp, QDnBTUn; ZymoMiPr1-3, ZymoMiPrSp, ZymoMIPrUn

### Key Phyloseq Objects

-   `DNAkits2`: Complete DNA kit comparison dataset (14 samples)
-   `DNAkits2.top20`: Top 20 taxa normalized to relative abundance
-   `DNAkits2.top20_family`: Top 20 families agglomerated at Family level

### Metadata Variables

-   `DNA Kit`: Extraction kit used (Presto Mini gDNA, Qiagen DNeasy Blood & Tissue, ZymoResearch DNA Miniprep)
-   `Sample`: Sample type (Chicken Rinse, Spiked, Unspiked)
-   `Number/Type`: Replicate number (1, 2, 3) or Control type
-   `Name`: Combined label (Number + Sample) for plotting

------------------------------------------------------------------------

## Visualization Specifications (All Datasets)

-   **Abundance transformations**:
    -   Relative abundance: Raw counts → proportions (0-1 scale)
    -   Percentage abundance: Raw counts → percentages (0-100 scale) via `transform_sample_counts()`
-   **Color schemes**:
    -   ColorBrewer "Paired" palette for top taxa
    -   "Other" (grouped low-abundance taxa) in grey85 (light grey)
    -   Custom pathogen-specific colors: Campylobacterota (#1b9e77 green), Enterobacteriaceae (#d95f02 orange)
-   **Taxonomic levels analyzed**: Phylum, Family, Genus, Species depending on analysis depth
-   **Top N taxa display**: Typically top 10-20 most abundant taxa, with remaining taxa grouped as "Other"
-   **Plot types**:
    -   Stacked bar plots: Taxonomic composition over time/treatment
    -   Alpha diversity plots: Shannon and Simpson indices
    -   Ordination plots: NMDS with Bray-Curtis dissimilarity
    -   Heatmaps: Read depth quality control, abundance patterns
    -   Combined plots: Using `patchwork`, `cowplot`, and `aplot` packages

------------------------------------------------------------------------

## Analysis Goals

This comprehensive workflow enables: 1. **Enrichment optimization**: Identify optimal enrichment broths and incubation times for pathogen detection 2. **Host DNA reduction**: Evaluate efficacy of depletion methods and sample processing approaches 3. **Extraction optimization**: Compare DNA extraction kit performance for food microbiome samples 4. **Pathogen tracking**: Monitor *Campylobacter* (Campylobacterota) and *Salmonella* (Enterobacteriaceae) enrichment dynamics 5. **Quality control**: Track read depths and sequencing quality across all experimental conditions 6. **Method validation**: Assess reproducibility across technical replicates and procedural blanks

------------------------------------------------------------------------

## Key R Packages Used

-   **DADA2**: Amplicon sequencing pipeline (denoising, chimera removal)
-   **phyloseq**: Microbiome data structure and analysis
-   **ggplot2**: Data visualization
-   **dplyr/tidyr**: Data manipulation
-   **stringr**: String pattern extraction
-   **RColorBrewer/scales**: Color palettes and axis formatting
-   **patchwork/cowplot/aplot**: Multi-panel plot composition
-   **pheatmap**: Heatmap generation

```{r Install Packages (if needed)}
install.packages("rmarkdown")

install.packages("BiocManager")

install.packages("pheatmap")

install.packages("aplot")

install.packages("tidyr")

install.packages("cowplot")

install.packages("stringr")

install.packages("dplyr")

BiocManager::install("dada2")

BiocManager::install("phyloseq")

```

```{r Load Packages & check version}
library(rmarkdown)

#Load dada2 and check version 
library(dada2)
packageVersion("dada2")

#Load ShortRead and check version 
library(ShortRead)
packageVersion("ShortRead")

#Load ggplot2 and check version 
library(ggplot2)
packageVersion("ggplot2")

#Load phyloseq
library(phyloseq)
packageVersion("phyloseq")

#Load stringr
library(stringr)
packageVersion("stringr")

#Load ggplot2
library(ggplot2)
packageVersion("ggplot2")

#Load dplyr
library(dplyr)
packageVersion("dplyr")

#Load pheatmap (for heat maps)
library(pheatmap)
packageVersion("pheatmap")

#Load aplot
library(aplot)
packageVersion("aplot")

#Load tidyr
library(tidyr)
packageVersion("tidyr")

#Load cowplot
library(cowplot)
packageVersion("cowplot")

#Load RColorBrewer
library(RColorBrewer)
packageVersion("RColorBrewer")

#Load RColorBrewer
library(scales)
packageVersion("scales")


```

#DADA2: Data Analysis of Sequence Data

```{r Set Path}
path <- "/nesi/nobackup/massey04083/method_development_chapter1/16S_Sequencing/MGS00721_Alice_Bradbury_Delivery/MGS00721_1/fQsequences"
path
```

```{r List files within the path}
fns <- list.files(path)
fns
```

```{r Define full path}
#Extract fastq sequences
fastqs <- fns[grepl(".fastq.gz", fns)]
fastqs

#Sort them to ensure reads are in the same order
fastqs <- sort(fastqs)

#Make sub-lists for the forward and reverse reads
fnFs <- fastqs[grepl("_R1", fastqs)] # Just the forward read files
fnRs <- fastqs[grepl("_R2", fastqs)] # Just the reverse read files

#Get the sample.names
sample.names <- sapply(strsplit(fnFs, "_"), `[`, 1)

#Specify the full path to the fnFs and fnRs
fnFs <- file.path(path, fnFs)
fnRs <- file.path(path, fnRs)

fnFs
fnRs

```

```{r Quality Plots}
#Visualize the quality profile of the forward reads
plotQualityProfile(fnFs[[1]])
plotQualityProfile(fnFs[[12]])
plotQualityProfile(fnRs[[1]])
plotQualityProfile(fnRs[[12]])

```

```{r Make directory for filtered files}
# Make directory and filenames for the filtered fastqs
filt_path <- file.path(path, "filtered")
if(!file_test("-d", filt_path)) dir.create(filt_path)
filt_path

# make list of filtered names for later
filtFs <- file.path(filt_path, paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(filt_path, paste0(sample.names, "_R_filt.fastq.gz"))

```

```{r Trimming and Filtering}
#Perform the trimming and filtering
#collect the names
names(filtFs) <- sample.names
names(filtRs) <- sample.names

#Perform the trimming and filtering
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, truncLen=c(230,200),
         maxN=0, maxEE=c(2,2), truncQ=2, rm.phix=TRUE,
         compress=TRUE, multithread=FALSE)
head(out)

```

```{r Dereplicate and rename}
#dereplicate the forward and reverse reads separately
derepFs <- derepFastq(filtFs, verbose=TRUE)

derepRs <- derepFastq(filtRs, verbose=TRUE)

##rename the derep-class objects by the sample names
names(derepFs) <- sample.names
names(derepRs) <- sample.names

names(derepFs)
```

```{r Learn errors for forward and reverse & plot these}
#Forward reads first and then look at the output
errF <- learnErrors(derepFs, multithread=FALSE)

dadaFs.lrn <- dada(derepFs, err=errF, multithread=TRUE)

#Reverse reads second and then look at the output
errR <- learnErrors(derepRs, multithread=FALSE)

dadaRs.lrn <- dada(derepRs, err=errR, multithread=TRUE)

##Plot the errors as a trellis plot

plotErrors(errF, nominalQ=TRUE)
```

```{r Sequence variants in each sample}
#Infer the sequence variants in each sample
dadaFs <- dada(derepFs, err=errF, multithread=TRUE)

dadaRs <- dada(derepRs, err=errR, multithread=TRUE)

#inspect the results in more detail
dadaFs[[1]]

dadaFs[[1]][10]$opts

dadaFs[[1]][10]$opts$BAND_SIZE

dadaFs[[1]][10]$opts$USE_QUALS

dadaFs[[1]][10]$opts$VECTORIZED_ALIGNMENT
```

```{r Merge denoised forward and reversed reads}
#Merge the denoised forward and reverse reads

mergers <- mergePairs(dadaFs, derepFs, dadaRs, derepRs, verbose=TRUE)

head(mergers[[1]])
```

```{r Sequence Table and Lengths}
#Construct sequence table of the non-Mock data

seqtab <- makeSequenceTable(mergers)

#Inspect the distribution of sequence lengths

table(nchar(getSequences(seqtab)))

```

```{r Remove Chimetic Sequences}
### Remove chimeric sequences
seqtab.nochim <- removeBimeraDenovo(seqtab, verbose=TRUE)

sum(seqtab.nochim)/sum(seqtab)

#Remove chimeric sequences with some complicated code
getN <- function(x) sum(getUniques(x)) # our first function
track <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN),
           sapply(mergers, getN), rowSums(seqtab.nochim))

colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged",
                       "nonchim")

rownames(track) <- sample.names

head(track)

```

```{r Assign Taxonomy to species level}
#use a reference data set to assign taxonomy to the reads

training <- ("~/00_nesi_projects/00_nesi_projects/massey04083_nobackup/method_development_chapter1/16S_Sequencing/silva_nr99_v138.2_toSpecies_trainset.fa")
training

taxa <- assignTaxonomy(
  seqtab.nochim,
  training,
  multithread = 12,
  verbose = TRUE
)
taxa

taxa.print <- taxa
rownames(taxa.print) <- NULL
head(taxa.print)
```

```{r OTU table}

samples.out <- rownames(seqtab.nochim)
samples.out

ps <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows=FALSE), sample_names(samples.out), tax_table(taxa))
ps

```

```{r All - Alpha Richness plot - Shannon & Simpson}

plot_richness(ps, measures=c("Shannon", "Simpson"))

```

```{r All - Bray NMDS}
## Ordination method
ord.nmds.bray <- ordinate(ps, method="NMDS", distance="bray")
plot_ordination(ps, ord.nmds.bray, title="Bray NMDS")
```

```{r All - Plot top 20}

# Agglomerate taxa at the Species level - this collapses all taxa that have the same species-level classification into single entries, summing their counts
ps_species_level_all96 <- tax_glom(ps, taxrank = "Genus")

# Identify the top 20 most abundant taxa across all 96 samples, sort in descending order and select the first 20 taxon names
top20_taxa_names_all96 <- names(sort(taxa_sums(ps_species_level_all96), decreasing=TRUE))[1:20]

# Display the first few taxon IDs from the top 20 list
head(top20_taxa_names_all96)

# Transform counts to relative abundances 
ps_relative_abundance_all96 <- transform_sample_counts(ps_species_level_all96, function(OTU) OTU/sum(OTU))

# Subset the phyloseq object to retain only the top 20 most abundant taxa, r
# Removes all other taxa
ps_top20_rel_abund_all96 <- prune_taxa(top20_taxa_names_all96, ps_relative_abundance_all96)

# Create a bar plot showing the relative abundance of the top 20 taxa across all 96 samples,
plot_top20_by_species_all96 <- plot_bar(ps_top20_rel_abund_all96, fill="Genus")

# Display plot 
plot_top20_by_species_all96


plot_df <- psmelt(ps_top20_rel_abund_all96)

# Replace NA genus with "Other" (CRITICAL)
plot_df$Genus <- as.character(plot_df$Genus)
plot_df$Genus[is.na(plot_df$Genus)] <- "Other"

# Get exact genus levels used in the plot
genus_levels <- sort(unique(plot_df$Genus))



library(dplyr)
library(stringr)
library(ggplot2)
library(RColorBrewer)

# ------------------------------------------------
# Build sample ordering metadata
# ------------------------------------------------
sample_info <- plot_df %>%
  distinct(Sample) %>%
  mutate(
    BrothGroup = case_when(
      str_detect(Sample, "^BFBB") ~ "BFBB",
      str_detect(Sample, "^CAT")  ~ "CAT",
      str_detect(Sample, "^BPW")  ~ "BPW",
      str_detect(Sample, "^RVS")  ~ "RVS",
      str_detect(Sample, "^NH")   ~ "NH",
      str_detect(Sample, "^H")    ~ "H",
      str_detect(Sample, "^HF|^NHCe") ~ "Filter",
      str_detect(Sample, "PrMig|QDnBT|Zymo|PRM") ~ "Extraction",
      str_detect(Sample, "ToxoVax|jejuni|Undetermined") ~ "Control",
      TRUE ~ "Other"
    ),

    Incubation = as.numeric(str_extract(Sample, "\\d+")),
    ND         = str_detect(Sample, "ND"),
    Replicate  = as.numeric(str_extract(Sample, "(?<=Ch|Ce)\\d"))
  ) %>%
  mutate(
    GroupOrder = case_when(
      BrothGroup %in% c("BFBB", "CAT", "BPW", "RVS") ~ 1,
      BrothGroup == "NH" ~ 2,
      BrothGroup == "H"  ~ 3,
      BrothGroup == "Filter" ~ 4,
      BrothGroup == "Extraction" ~ 5,
      BrothGroup == "Control" ~ 6,
      TRUE ~ 99
    )
  )

sample_order <- sample_info %>%
  arrange(
    GroupOrder,
    BrothGroup,
    Incubation,
    !ND,
    Replicate,
    Sample
  ) %>%
  pull(Sample)

plot_df <- plot_df %>%
  mutate(Sample = factor(Sample, levels = sample_order))

# ------------------------------------------------
# Colour palette
# ------------------------------------------------
genus_levels <- levels(factor(plot_df$Genus))

pal_genera <- colorRampPalette(
  brewer.pal(12, "Paired")
)(length(genus_levels))

names(pal_genera) <- genus_levels

if ("Other" %in% names(pal_genera)) {
  pal_genera["Other"] <- "grey85"
}

# ------------------------------------------------
# Plot
# ------------------------------------------------
label_lines <- plot_df %>%
  distinct(Sample)


plot_top20_by_species_all96 <- ggplot(
  plot_df,
  aes(x = Sample, y = Abundance, fill = Genus)
) +
  geom_bar(stat = "identity", width = 0.9) +

  scale_fill_manual(values = pal_genera, drop = FALSE) +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +

  theme_bw(base_size = 16) +
  theme(
    ## 🔹 REMOVE black border only
    panel.border = element_blank(),

    ## 🔹 vertical guide lines aligned with bars
    panel.grid.major.x = element_line(
      colour = "grey85",
      linewidth = 0.3
    ),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),

    ## Labels directly under bars
    axis.text.x = element_text(
      size = 10,
      angle = 90,
      hjust = 1,
      vjust = 0.5
    ),

    axis.text.y  = element_text(size = 16),
    axis.title.x = element_text(size = 16, face = "bold"),
    axis.title.y = element_text(size = 16, face = "bold"),
    legend.text  = element_text(size = 14),
    legend.title = element_text(size = 16, face = "bold"),

    plot.margin = margin(10, 10, 10, 10)
  ) +
  labs(
    x = "Sample",
    y = "Relative abundance",
    fill = "Genus"
  )

# Display
plot_top20_by_species_all96


# -----------------------------------
# Save
# -----------------------------------

ggsave("taxa20_all96samples.png", 
       plot=plot_top20_by_species_all96,
       width=16,
       height=10,
       dpi=600)
```


```{r Top 20 - including all bacteria as other - all 96 samples}

library(phyloseq)
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(stringr)
library(scales)

# ---- 1) Agglomerate at Genus level ----
ps_genus <- tax_glom(ps, taxrank = "Genus")

# ---- 2) Transform counts to relative abundance ----
ps_rel <- transform_sample_counts(ps_genus, function(x) x / sum(x))

# ---- 3) Melt to long dataframe ----
df <- psmelt(ps_rel) %>%
  mutate(
    Genus = ifelse(is.na(Genus) | Genus == "", "Unassigned", as.character(Genus))
  )

# ---- 4) Identify top 20 genera across all samples ----
top20_genera <- df %>%
  group_by(Genus) %>%
  summarise(total_abundance = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_abundance)) %>%
  slice_head(n = 20) %>%
  pull(Genus)

# ---- 5) Assign "Other" to genera not in top 20 ----
df <- df %>%
  mutate(
    Genus2 = ifelse(Genus %in% top20_genera, Genus, "Other")
  )

# ---- 6) Set factor levels: top genera first, then Unassigned, then Other ----
genus_order <- df %>%
  group_by(Genus2) %>%
  summarise(total = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total)) %>%
  pull(Genus2)

# Put "Unassigned" and "Other" at the end
genus_order <- c(setdiff(genus_order, c("Unassigned", "Other")), "Unassigned", "Other")
df$Genus2 <- factor(df$Genus2, levels = genus_order)

# ---- 7) Colour palette ----
pal <- colorRampPalette(brewer.pal(12, "Paired"))(length(levels(df$Genus2)))
names(pal) <- levels(df$Genus2)
pal["Other"] <- "grey85"
pal["Unassigned"] <- "grey65"

# ---- 8) Optional: build sample ordering (adapt your previous logic) ----
sample_info <- df %>%
  distinct(Sample) %>%
  mutate(
    BrothGroup = case_when(
      str_detect(Sample, "^BFBB") ~ "BFBB",
      str_detect(Sample, "^CAT")  ~ "CAT",
      str_detect(Sample, "^BPW")  ~ "BPW",
      str_detect(Sample, "^RVS")  ~ "RVS",
      str_detect(Sample, "^NH")   ~ "NH",
      str_detect(Sample, "^H")    ~ "H",
      str_detect(Sample, "^HF|^NHCe") ~ "Filter",
      str_detect(Sample, "PrMig|QDnBT|Zymo|PRM") ~ "Extraction",
      str_detect(Sample, "ToxoVax|jejuni|Undetermined") ~ "Control",
      TRUE ~ "Other"
    ),
    Incubation = as.numeric(str_extract(Sample, "\\d+")),
    ND = str_detect(Sample, "ND"),
    Replicate = as.numeric(str_extract(Sample, "(?<=Ch|Ce)\\d"))
  ) %>%
  mutate(
    GroupOrder = case_when(
      BrothGroup %in% c("BFBB", "CAT", "BPW", "RVS") ~ 1,
      BrothGroup == "NH" ~ 2,
      BrothGroup == "H"  ~ 3,
      BrothGroup == "Filter" ~ 4,
      BrothGroup == "Extraction" ~ 5,
      BrothGroup == "Control" ~ 6,
      TRUE ~ 99
    )
  )

sample_order <- sample_info %>%
  arrange(
    GroupOrder,
    BrothGroup,
    Incubation,
    !ND,
    Replicate,
    Sample
  ) %>%
  pull(Sample)

df$Sample <- factor(df$Sample, levels = sample_order)

# ---- 9) Plot stacked bar chart ----
plot_all_top20 <- ggplot(df, aes(x = Sample, y = Abundance, fill = Genus2)) +
  geom_bar(stat = "identity", width = 0.9) +
  scale_fill_manual(values = pal, drop = FALSE) +
  scale_x_discrete(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0), labels = scales::percent_format()) +
  theme_bw(base_size = 16) +
  theme(
    panel.border = element_blank(),
    panel.grid.major.x = element_line(colour = "grey85", linewidth = 0.3),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    axis.text.x = element_text(size = 10, angle = 90, hjust = 1, vjust = 0.5),
    axis.text.y = element_text(size = 16),
    axis.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 16, face = "bold"),
    plot.margin = margin(10,10,10,10)
  ) +
  labs(
    x = "Sample",
    y = "Relative abundance",
    fill = "Genus"
  ) +
  guides(fill = guide_legend(ncol = 1))  # <-- forces legend into a single column

plot_all_top20


# ---- 10) Save ----
ggsave("taxa_top20_all_samples_with-other.png", plot = plot_all_top20,
       width = 18, height = 10, dpi = 600)


```



#Read Depths and Abundance - Heat Map

```{r Combined Abundance Bar Plot and Read Depth Heat Map}

sample_order <- levels(plot_df$Sample)

# Prepare heatmap data from track
heatmap_df <- track %>%
  as.data.frame() %>%
  rownames_to_column("Sample") %>%
  pivot_longer(
    cols = -Sample,
    names_to = "Step",
    values_to = "Reads"
  )

# Ensure correct DADA2 step order
heatmap_df$Step <- factor(
  heatmap_df$Step,
  levels = c(
    "input",
    "filtered",
    "denoisedF",
    "denoisedR",
    "merged",
    "nonchim"
  )
)

heatmap_df <- heatmap_df %>%
  mutate(
    Sample = factor(Sample, levels = sample_order)
  )

# Build heatmap
heatmap_reads <- ggplot(
  heatmap_df,
  aes(x = Step, y = Sample, fill = Reads)
) +
  geom_tile(color = "grey90", linewidth = 0.15) +

  scale_fill_viridis_c(
    option = "magma",
    trans  = "log10",
    name   = "Reads (log10)"
  ) +

  theme_minimal(base_size = 12) +
  theme(
    axis.text.y  = element_text(size = 8),
    axis.text.x  = element_text(size = 12, angle = 45, hjust = 1),
    axis.title.y = element_text(size = 14, face = "bold"),
    axis.title.x = element_text(size = 14, face = "bold"),
    panel.grid   = element_blank(),
    plot.margin  = margin(10, 10, 10, 10)
  ) +

  labs(
    title = "Read retention across DADA2 pipeline steps",
    x = "Pipeline step",
    y = "Sample"
  )

# Display
heatmap_reads

# Save
ggsave(
  "read_counts_changes_through_DADA2.png",
  plot   = heatmap_reads,
  width  = 8,
  height = 14,
  dpi    = 600
)

```

```{r Combined Bar Plot and Heatmap}

# Combine the bar plot and heatmap vertically using cowplot
combined_plot_cowplot <- plot_grid(
  plot_abundance_families,                                           # Bar plot (top)
  plot_heatmap_reads,                                                # Heatmap (bottom)
  ncol = 1,                                                          # Stack vertically (1 column)
  rel_heights = c(4, 1),                                             # 4:1 height ratio
  align = 'v'                                                        # Align vertically
)

# Display the combined plot
combined_plot_cowplot

# Save the cowplot version
ggsave("combined_abundance_reads_cowplot.png", 
       plot = combined_plot_cowplot, 
       width = 12, 
       height = 8, 
       dpi = 300)
```

#Enrichment testing analysis

```{r Enrichment: Data processing}
# Display all sample names available in the dataset
samples.out

# Define the original sample codes for enrichment testing conditions
# Format: [Medium][Timepoint] where medium = BFBB/CAT/BPW/RVS and timepoint = 0/4/8/12/24 hours
enrichment_testing_codes <- c("BFBB0", "BFBB4", "BFBB8", "BFBB12", "BFBB24", 
                               "CAT0", "CAT4", "CAT8", "CAT12", "CAT24", 
                               "BPW0", "BPW4", "BPW8", "BPW12", "BPW24", 
                               "RVS0", "RVS4", "RVS8", "RVS12", "RVS24")

# Display the enrichment testing sample codes
enrichment_testing_codes

# Filter to keep only enrichment testing samples that exist in the dataset
enrichment_testing_samples <- enrichment_testing_codes[enrichment_testing_codes %in% samples.out]

# Display the filtered enrichment testing samples
enrichment_testing_samples

# Replace medium code "BFBB" with full descriptive name "BF Bolton base "
enrichment_testing_samples <- gsub("BFBB", "BF Bolton base ", enrichment_testing_samples)

# Replace medium code "CAT" with full descriptive name "BF Bolton base w CAT "
enrichment_testing_samples <- gsub("CAT", "BF Bolton base plus CAT ", enrichment_testing_samples)

# Replace medium code "RVS" with full descriptive name "RVS "
enrichment_testing_samples <- gsub("RVS", "RVS ", enrichment_testing_samples)

# Replace medium code "BPW" with full descriptive name "BPW "
enrichment_testing_samples <- gsub("BPW", "BPW ", enrichment_testing_samples)

# Replace timepoint "0" with "0 hours"
enrichment_testing_samples <- gsub("0", "0 hours", enrichment_testing_samples)

# Replace timepoint "8" with "8 hours"
enrichment_testing_samples <- gsub("8", "8 hours", enrichment_testing_samples)

# Replace timepoint "12" with "12 hours"
enrichment_testing_samples <- gsub("12", "12 hours", enrichment_testing_samples)

# Replace timepoint "4" with "4 hours" (fixed=TRUE ensures exact match, 
# avoiding partial matches with "24")
enrichment_testing_samples <- gsub("4", "4 hours", enrichment_testing_samples, fixed=TRUE)

# Display the final enrichment testing sample names
enrichment_testing_samples
```

```{r Enrichment: Create data frame }

# Extract the broth types from the sample names using regex - [A-Za-z ]+ captures all letters and spaces (the broth names)
enrichment_broth_types <- str_extract(enrichment_testing_samples, "[A-Za-z ]+")
enrichment_broth_types

# Extract the incubation time (numeric values only) from the sample names using regex - \\d+ captures one or more digits 
enrichment_incubation_times <- str_extract(enrichment_testing_samples, "\\d+")
enrichment_incubation_times

# Create a sample metadata dataframe with two columns: Broth (medium type) and Incubation (time in hours)
samdf_enrichment <- data.frame(Broth=enrichment_broth_types, 
                                Incubation=enrichment_incubation_times)

# Display dataframe
samdf_enrichment

# Set row names to the original sample codes (BFBB0, CAT4, etc.) for matching with the OTU table
rownames(samdf_enrichment) <- enrichment_testing_codes

# Display the dataframe with row names
samdf_enrichment

# Create a phyloseq object specifically for enrichment testing samples
# Combines:  OTU abundance table, sample metadata, and taxonomic classifications
ps_enrichment_testing <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows=FALSE), 
                                   sample_data(samdf_enrichment), 
                                   tax_table(taxa))

# Display summary of the enrichment testing phyloseq object
ps_enrichment_testing
```

##Enrichment Testing - Alpha Diversity Analysis

```{r Enrichment: FIGURE 1-Alpha diversity}

# Extract sample metadata from phyloseq object as a dataframe
sample_metadata_enrichment <- as.data.frame(sample_data(ps_enrichment_testing))

# Remove leading/trailing whitespace from Broth column
sample_metadata_enrichment$Broth <- trimws(as.character(sample_metadata_enrichment$Broth))

# Create numeric incubation time variable - Converts "4 hours" → 4, "24 hours" → 24, etc.
sample_metadata_enrichment$Incubation_hr <- as.numeric(gsub("[^0-9.]", "", as.character(sample_metadata_enrichment$Incubation)))

# Update the phyloseq object with cleaned Broth values
sample_data(ps_enrichment_testing)$Broth <- sample_metadata_enrichment$Broth

# Add the new numeric incubation time variable to the phyloseq object
sample_data(ps_enrichment_testing)$Incubation_hr <- sample_metadata_enrichment$Incubation_hr

#Create alpha diversity plot (Shannon and Simpson indices) over incubation time

plot_alpha_diversity_enrichment <- plot_richness(
  ps_enrichment_testing,
  x = "Incubation_hr",                              # X-axis:  incubation time in hours
  measures = c("Shannon", "Simpson")                # Diversity metrics to calculate
) +
  aes(color = Broth) +                              # Color points by broth/medium type
  scale_color_manual(values = c(                    # Custom color palette for each broth
    "BF Bolton base"       = "#F8766D",             # Red for BF Bolton base
    "BF Bolton base plus CAT" = "#7CAE00",             # Cyan for BF Bolton base with CAT
    "BPW"                  = "#00BFC4",             # Green for Buffered Peptone Water
    "RVS"                  = "#C77CFF"              # Purple for Rappaport-Vassiliadis Soya
  )) +
  scale_x_continuous(breaks = c(0, 4, 8, 12, 24)) + # Set X-axis breaks at each timepoint
  labs(x = "Incubation (h)")   +                      # X-axis label
  geom_point(size = 3, alpha = 0.6)

# Display the alpha diversity plot
plot_alpha_diversity_enrichment

ggsave("alpha_diversity_enrichment.png", 
       plot = plot_alpha_diversity_enrichment, 
       width = 10, 
       height = 6, 
       dpi = 600)

```

##Enrichment Testing - Bray NMDS

```{r Enrichment: Bray NMDS}

# Perform Non-metric Multidimensional Scaling (NMDS) ordination using Bray-Curtis dissimilarity
ordination_nmds_bray_enrichment <- ordinate(ps_enrichment_testing, 
                                             method = "NMDS",      # NMDS ordination method
                                             distance = "bray")    # Bray-Curtis dissimilarity metric

# Create ordination plot 
plot_ordination_bray_enrichment <- plot_ordination(ps_enrichment_testing, 
                                                    ordination_nmds_bray_enrichment, 
                                                    title = "Bray NMDS")

```

##Enrichment Testing: Plots - Microbial Diversity

```{r Combined Plot Using Aplot - top 20 Families & reads}

library(phyloseq)
library(dplyr)
library(tidyr)
library(ggplot2)
library(tibble)

# Alpha diversity
div_df <- phyloseq::estimate_richness(
  ps_top20_families_enrichment,
  measures = "Shannon"
)
div_df$Sample <- rownames(div_df)

# Sample metadata
meta_df <- as.data.frame(phyloseq::sample_data(ps_top20_families_enrichment))
meta_df$Sample <- rownames(meta_df)

# Read counts
reads_df <- data.frame(
  Sample = phyloseq::sample_names(ps_top20_families_enrichment),
  Reads  = phyloseq::sample_sums(ps_top20_families_enrichment)
)

# Combine into ONE dataframe
heatmap_df <- div_df %>%
  dplyr::left_join(meta_df, by = "Sample") %>%
  dplyr::left_join(reads_df, by = "Sample")

head(heatmap_df)


```

##All Enrichment Broths - Top 12 Genera Stacked Bar Plot

```{r Enrichment Broths All: Top 12 Genera Stacked Bar Plot (FIXED)}

library(phyloseq)
library(dplyr)
library(ggplot2)
library(scales)
library(RColorBrewer)
library(stringr)


# Extract and clean sample metadata
sample_metadata_all_broths <- as.data.frame(sample_data(ps_enrichment_testing))
sample_metadata_all_broths$Broth <- trimws(as.character(sample_metadata_all_broths$Broth))
sample_metadata_all_broths$Incubation_hr <- as.numeric(gsub("[^0-9.]", "", as.character(sample_metadata_all_broths$Incubation)))

# Update phyloseq object with cleaned metadata
sample_data(ps_enrichment_testing)$Broth <- sample_metadata_all_broths$Broth
sample_data(ps_enrichment_testing)$Incubation_hr <- sample_metadata_all_broths$Incubation_hr

# Agglomerate at Genus level
ps_genus_all_broths <- tax_glom(ps_enrichment_testing, taxrank = "Genus", NArm = FALSE)
ps_genus_all_broths <- prune_taxa(taxa_sums(ps_genus_all_broths) > 0, ps_genus_all_broths)

# Transform to relative abundance
ps_genus_rel_all_broths <- transform_sample_counts(ps_genus_all_broths, function(x) x / sum(x))

# Convert to long format and clean taxonomy
df_all_broths <- psmelt(ps_genus_rel_all_broths) %>%
  mutate(
    Broth = trimws(as.character(Broth)),
    Incubation_hr = as.numeric(Incubation_hr),
    Genus = ifelse(is.na(Genus) | Genus == "", "Unassigned", as.character(Genus))
  )

# Identify top 12 genera across all enrichment broths
topN <- 20
top_genera_all_broths <- df_all_broths %>%
  group_by(Genus) %>%
  summarise(tot = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(tot)) %>%
  slice_head(n = topN) %>%
  pull(Genus)

# Top genera as themselves, everything else = Other
df_all_broths <- df_all_broths %>%
  mutate(Genus2 = ifelse(Genus %in% top_genera_all_broths, Genus, "Other"))

# Order legend by total abundance (Other last)
genus_order_all <- df_all_broths %>%
  group_by(Genus2) %>%
  summarise(tot = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(tot)) %>%
  pull(Genus2)
genus_order_all <- c(setdiff(genus_order_all, "Other"), "Other")
df_all_broths$Genus2 <- factor(df_all_broths$Genus2, levels = genus_order_all)

# Palette (grey for Other)
n_genera_all <- length(levels(df_all_broths$Genus2))
pal_genera_all <- colorRampPalette(brewer.pal(20, "Paired"))(n_genera_all)
names(pal_genera_all) <- levels(df_all_broths$Genus2)
pal_genera_all["Other"] <- "grey85"
pal_genera_all

# Plot: faceted stacked bars
plot_top_genera_all_broths <- df_all_broths %>%
  group_by(Broth, Sample, Incubation_hr, Genus2) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop") %>%
  ggplot(aes(x = Incubation_hr, y = Abundance, fill = Genus2)) +
  geom_col(width = 2, color = "white", linewidth = 0.25) +
  facet_wrap(~Broth, ncol = 2) +
  scale_x_continuous(breaks = c(0, 4, 8, 12, 24)) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = pal_genera_all, drop = FALSE) +
    guides(fill = guide_legend(ncol = 1)) + 
  labs(x = "Incubation (h)", y = "Relative abundance", fill = "Genus", title =
    "Relative abundance of top 20 genera for enrichment broths during incubation") +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    legend.key.height = unit(0.45, "cm")
  )

plot_top_genera_all_broths

ggsave("plot_top_genera_all_broths.png", plot = plot_top_genera_all_broths, width = 12, height = 8, dpi = 600)


```

```{r Heat Map - top 20 enrichment testing}
# Load required libraries
library(phyloseq)
library(tidyverse)
library(RColorBrewer)
library(scales)
library(ComplexHeatmap)
BiocManager::install("ComplexHeatmap")
library(circlize)

# Extract and clean sample metadata
sample_metadata_all_broths <- as.data.frame(sample_data(ps_enrichment_testing))
sample_metadata_all_broths$Broth <- trimws(as.character(sample_metadata_all_broths$Broth))
sample_metadata_all_broths$Incubation_hr <- as.numeric(gsub("[^0-9.]", "", as.character(sample_metadata_all_broths$Incubation)))

# Update phyloseq object with cleaned metadata
sample_data(ps_enrichment_testing)$Broth <- sample_metadata_all_broths$Broth
sample_data(ps_enrichment_testing)$Incubation_hr <- sample_metadata_all_broths$Incubation_hr

# Agglomerate at Genus level
ps_genus_all_broths <- tax_glom(ps_enrichment_testing, taxrank = "Genus", NArm = FALSE)
ps_genus_all_broths <- prune_taxa(taxa_sums(ps_genus_all_broths) > 0, ps_genus_all_broths)

# Get absolute counts (for annotation)
ps_genus_counts <- ps_genus_all_broths

# Transform to relative abundance
ps_genus_rel_all_broths <- transform_sample_counts(ps_genus_all_broths, function(x) x / sum(x))

# Convert to long format and clean taxonomy - RELATIVE ABUNDANCE
df_all_broths <- psmelt(ps_genus_rel_all_broths) %>%
  mutate(
    Broth = trimws(as.character(Broth)),
    Incubation_hr = as.numeric(Incubation_hr),
    Genus = ifelse(is.na(Genus) | Genus == "", "Unassigned", as.character(Genus))
  )

# Convert to long format - ABSOLUTE COUNTS
df_counts <- psmelt(ps_genus_counts) %>%
  mutate(
    Broth = trimws(as.character(Broth)),
    Incubation_hr = as.numeric(Incubation_hr),
    Genus = ifelse(is.na(Genus) | Genus == "", "Unassigned", as.character(Genus))
  )

# Identify top 20 genera across all enrichment broths
topN <- 20
top_genera_all_broths <- df_all_broths %>%
  group_by(Genus) %>%
  summarise(tot = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(tot)) %>%
  slice_head(n = topN) %>%
  pull(Genus)

# Filter for top genera only
df_top_rel <- df_all_broths %>%
  filter(Genus %in% top_genera_all_broths)

df_top_counts <- df_counts %>%
  filter(Genus %in% top_genera_all_broths)

# Create matrix for heatmap - RELATIVE ABUNDANCE
# Average across replicates if needed
mat_rel <- df_top_rel %>%
  group_by(Broth, Incubation_hr, Genus) %>%
  summarise(Mean_Abundance = mean(Abundance, na.rm = TRUE), .groups = "drop") %>%
  mutate(Sample_ID = paste(Broth, Incubation_hr, "h", sep = "_")) %>%
  select(Genus, Sample_ID, Mean_Abundance) %>%
  pivot_wider(names_from = Sample_ID, values_from = Mean_Abundance, values_fill = 0) %>%
  column_to_rownames("Genus") %>%
  as.matrix()

# Create matrix for heatmap - READ COUNTS
mat_counts <- df_top_counts %>%
  group_by(Broth, Incubation_hr, Genus) %>%
  summarise(Mean_Counts = mean(Abundance, na.rm = TRUE), .groups = "drop") %>%
  mutate(Sample_ID = paste(Broth, Incubation_hr, "h", sep = "_")) %>%
  select(Genus, Sample_ID, Mean_Counts) %>%
  pivot_wider(names_from = Sample_ID, values_from = Mean_Counts, values_fill = 0) %>%
  column_to_rownames("Genus") %>%
  as.matrix()

# Ensure both matrices have same dimensions and order
common_samples <- intersect(colnames(mat_rel), colnames(mat_counts))
common_genera <- intersect(rownames(mat_rel), rownames(mat_counts))
mat_rel <- mat_rel[common_genera, common_samples]
mat_counts <- mat_counts[common_genera, common_samples]

# Order genera by total abundance
genus_order <- rownames(mat_rel)[order(rowSums(mat_rel), decreasing = TRUE)]
mat_rel <- mat_rel[genus_order, ]
mat_counts <- mat_counts[genus_order, ]

# Create column annotation for broth type and time
col_anno_df <- data.frame(
  Sample = colnames(mat_rel),
  stringsAsFactors = FALSE
) %>%
  mutate(
    Broth = str_replace(Sample, "_\\d+_h$", ""),
    Time = str_extract(Sample, "\\d+(?=_h$)")
  )

# Create color functions
col_fun_rel <- colorRamp2(
  c(0, quantile(mat_rel, 0.5), max(mat_rel)),
  c("white", "yellow", "red")
)

col_fun_counts <- colorRamp2(
  c(0, quantile(mat_counts[mat_counts > 0], 0.5, na.rm = TRUE), max(mat_counts)),
  c("white", "lightblue", "darkblue")
)

# Column annotation colors
broth_colors <- setNames(
  brewer.pal(length(unique(col_anno_df$Broth)), "Set2"),
  unique(col_anno_df$Broth)
)

time_colors <- setNames(
  colorRampPalette(brewer.pal(9, "YlOrRd"))(length(unique(col_anno_df$Time))),
  sort(unique(col_anno_df$Time))
)

# Column annotations
ha_col <- HeatmapAnnotation(
  Broth = col_anno_df$Broth,
  Time = col_anno_df$Time,
  col = list(
    Broth = broth_colors,
    Time = time_colors
  ),
  annotation_name_side = "left",
  annotation_legend_param = list(
    Broth = list(title = "Broth Type"),
    Time = list(title = "Time (h)")
  )
)

# Create cell annotation function to show read counts
cell_fun <- function(j, i, x, y, width, height, fill) {
  if(mat_counts[i, j] > 0) {
    count_val <- round(mat_counts[i, j])
    if(count_val >= 100) {
      grid.text(sprintf("%.0f", count_val), x, y, 
                gp = gpar(fontsize = 6, col = "black"))
    }
  }
}

# VERSION 1: Relative abundance heatmap with read count annotations
ht_rel_annotated <- Heatmap(
  mat_rel * 100,  # Convert to percentage
  name = "Relative\nAbundance (%)",
  col = col_fun_rel,
  
  # Clustering
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  
  # Column settings
  column_title = "Enrichment Broths and Incubation Times",
  column_names_rot = 45,
  column_names_gp = gpar(fontsize = 8),
  top_annotation = ha_col,
  
  # Row settings
  row_title = "Genus",
  row_names_gp = gpar(fontsize = 9, fontface = "italic"),
  
  # Cell annotations
  cell_fun = cell_fun,
  
  # Heatmap appearance
  border = TRUE,
  rect_gp = gpar(col = "white", lwd = 1),
  
  # Legend
  heatmap_legend_param = list(
    title = "Relative\nAbundance (%)",
    at = c(0, 25, 50, 75, 100),
    labels = c("0", "25", "50", "75", "100"),
    direction = "vertical"
  )
)

# VERSION 2: Side-by-side heatmaps (relative abundance + read counts)
ht_rel <- Heatmap(
  mat_rel * 100,
  name = "Rel. Ab. (%)",
  col = col_fun_rel,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  column_title = "Relative Abundance",
  column_names_rot = 45,
  column_names_gp = gpar(fontsize = 8),
  row_names_gp = gpar(fontsize = 9, fontface = "italic"),
  top_annotation = ha_col,
  show_row_names = TRUE,
  border = TRUE,
  rect_gp = gpar(col = "white", lwd = 1),
  heatmap_legend_param = list(
    title = "Rel. Ab. (%)",
    direction = "vertical"
  )
)

ht_counts <- Heatmap(
  mat_counts,
  name = "Read Counts",
  col = col_fun_counts,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  column_title = "Read Counts",
  column_names_rot = 45,
  column_names_gp = gpar(fontsize = 8),
  row_names_gp = gpar(fontsize = 9, fontface = "italic"),
  show_row_names = FALSE,
  border = TRUE,
  rect_gp = gpar(col = "white", lwd = 1),
  heatmap_legend_param = list(
    title = "Read Counts",
    direction = "vertical"
  )
)

# VERSION 3: Log-scaled relative abundance with read count text
mat_rel_log <- log10(mat_rel * 100 + 0.1)  # Add pseudocount before log

cell_fun_text <- function(j, i, x, y, width, height, fill) {
  count_val <- round(mat_counts[i, j])
  if(count_val > 0) {
    grid.text(sprintf("%d", count_val), x, y, 
              gp = gpar(fontsize = 5, col = ifelse(mat_rel_log[i, j] > median(mat_rel_log), "white", "black")))
  }
}

ht_log_annotated <- Heatmap(
  mat_rel_log,
  name = "Log10(Rel. Ab.%)",
  col = colorRamp2(
    c(min(mat_rel_log), median(mat_rel_log), max(mat_rel_log)),
    c("white", "orange", "darkred")
  ),
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  column_title = "Log10 Relative Abundance (%) with Read Counts",
  column_names_rot = 45,
  column_names_gp = gpar(fontsize = 8),
  top_annotation = ha_col,
  row_names_gp = gpar(fontsize = 9, fontface = "italic"),
  cell_fun = cell_fun_text,
  border = TRUE,
  rect_gp = gpar(col = "white", lwd = 1),
  heatmap_legend_param = list(
    title = "Log10(Rel. Ab.%)",
    direction = "vertical"
  )
)

# Save plots
png("heatmap_genera_annotated.png", width = 14, height = 10, units = "in", res = 600)
draw(ht_rel_annotated, 
     column_title = "Top 20 Genera Across Enrichment Broths\nRelative Abundance (%) with Read Counts",
     padding = unit(c(2, 2, 2, 10), "mm"))
dev.off()

png("heatmap_genera_sidebyside.png", width = 16, height = 10, units = "in", res = 600)
draw(ht_rel + ht_counts, 
     column_title = "Top 20 Genera Across Enrichment Broths",
     padding = unit(c(2, 2, 2, 10), "mm"))
dev.off()

png("heatmap_genera_log_annotated.png", width = 14, height = 10, units = "in", res = 600)
draw(ht_log_annotated,
     padding = unit(c(2, 2, 2, 10), "mm"))
dev.off()

# Optional: Create a summary table with both metrics
summary_table <- df_top_rel %>%
  left_join(
    df_top_counts %>% 
      select(Broth, Incubation_hr, Genus, ReadCounts = Abundance),
    by = c("Broth", "Incubation_hr", "Genus")
  ) %>%
  group_by(Broth, Incubation_hr, Genus) %>%
  summarise(
    Mean_RelAbundance = mean(Abundance, na.rm = TRUE) * 100,
    Mean_ReadCounts = mean(ReadCounts, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(Broth, Incubation_hr, desc(Mean_RelAbundance))

write.csv(summary_table, "genera_abundance_summary.csv", row.names = FALSE)

# Print summary
cat("\nTop genera by total relative abundance:\n")
print(
  df_top_rel %>%
    group_by(Genus) %>%
    summarise(Total_RA = sum(Abundance) * 100, .groups = "drop") %>%
    arrange(desc(Total_RA)) %>%
    head(10)
)

cat("\nHeatmaps saved:\n")
cat("1. heatmap_genera_annotated.png - Relative abundance with read counts overlaid\n")
cat("2. heatmap_genera_sidebyside.png - Side-by-side comparison\n")
cat("3. heatmap_genera_log_annotated.png - Log-scaled with read counts\n")
cat("4. genera_abundance_summary.csv - Summary table exported\n")

```



```{r}
# -----------------------------------
# Agglomerate at Family level
# -----------------------------------

ps_Family_all_broths <- tax_glom(
  ps_enrichment_testing,
  taxrank = "Family",
  NArm = FALSE
)

ps_Family_all_broths <- prune_taxa(
  taxa_sums(ps_Family_all_broths) > 0,
  ps_Family_all_broths
)

# -----------------------------------
# Transform to relative abundance
# -----------------------------------

ps_Family_rel_all_broths <- transform_sample_counts(
  ps_Family_all_broths,
  function(x) x / sum(x)
)

# -----------------------------------
# Convert to long format & clean taxonomy
# -----------------------------------

df_all_broths_Family <- psmelt(ps_Family_rel_all_broths) %>%
  mutate(
    Broth = trimws(as.character(Broth)),
    Incubation_hr = as.numeric(Incubation_hr),
    Family = ifelse(
      is.na(Family) | Family == "",
      "Unassigned",
      as.character(Family)
    )
  )

# -----------------------------------
# Identify top 12 Families overall
# -----------------------------------

topN <- 20

top_Family_all_broths <- df_all_broths_Family %>%
  group_by(Family) %>%
  summarise(
    tot = sum(Abundance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(tot)) %>%
  slice_head(n = topN) %>%
  pull(Family)

# -----------------------------------
# Collapse non-top Families into "Other"
# -----------------------------------

df_all_broths_Family <- df_all_broths_Family %>%
  mutate(
    Family = ifelse(
      Family %in% top_Family_all_broths,
      Family,
      "Other"
    )
  )

# -----------------------------------
# Order legend by total abundance (Other last)
# -----------------------------------

Family_order_all <- df_all_broths_Family %>%
  group_by(Family) %>%
  summarise(
    tot = sum(Abundance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(tot)) %>%
  pull(Family)

Family_order_all <- c(setdiff(Family_order_all, "Other"), "Other")

df_all_broths_Family$Family <- factor(
  df_all_broths_Family$Family,
  levels = Family_order_all
)

# -----------------------------------
# Palette (Paired, grey for Other)
# -----------------------------------

n_family <- length(levels(df_all_broths_Family$Family))

pal_family_all <- colorRampPalette(
  brewer.pal(20, "Paired")
)(n_family)

names(pal_family_all) <- levels(df_all_broths_Family$Family)
pal_family_all["Other"] <- "grey85"

pal_family_all

# -----------------------------------
# Plot: faceted stacked bars (Family)
# -----------------------------------

plot_top_family_all_broths <- df_all_broths_Family %>%
  group_by(Broth, Sample, Incubation_hr, Family) %>%
  summarise(
    Abundance = sum(Abundance),
    .groups = "drop"
  ) %>%
  ggplot(
    aes(
      x = Incubation_hr,
      y = Abundance,
      fill = Family
    )
  ) +
  geom_col(
    width = 2,
    color = "white",
    linewidth = 0.25
  ) +
  facet_wrap(~ Broth, ncol = 2) +
  scale_x_continuous(breaks = c(0, 4, 8, 12, 24)) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = pal_family_all, drop = FALSE) +
    guides(fill = guide_legend(ncol = 1)) +
  labs(
    x = "Incubation (h)",
    y = "Relative abundance",
    fill = "Family",
    title = 
      "Relative abundance of top 20 families for enrichment broths during incubation"
    ) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    legend.key.height = unit(0.45, "cm")
  )

# Display
plot_top_family_all_broths

# Save
ggsave(
  "plot_top_family_all_broths.png",
  plot  = plot_top_family_all_broths,
  width = 12,
  height = 8,
  dpi   = 600
)

```

###Target Pathogen Analysis - How much Campylobacter and Salmonella in broths?

```{r ALl broths: Campylobacter and Salmoenlla present vs. All other bacteria}
# ===== CAMPYLOBACTER AND SALMONELLA FOCUSED GRAPHS - ALL ENRICHMENT BROTHS =====

# Extract and clean sample metadata
sample_metadata_pathogens <- as.data.frame(sample_data(ps_enrichment_testing))
sample_metadata_pathogens$Broth <- trimws(as.character(sample_metadata_pathogens$Broth))
sample_metadata_pathogens$Incubation_hr <- as.numeric(gsub("[^0-9.]", "", as.character(sample_metadata_pathogens$Incubation)))

# Update phyloseq object with cleaned metadata
sample_data(ps_enrichment_testing)$Broth <- sample_metadata_pathogens$Broth
sample_data(ps_enrichment_testing)$Incubation_hr <- sample_metadata_pathogens$Incubation_hr

# Subset to samples with valid metadata
ps_all_broths <- subset_samples(ps_enrichment_testing, ! is.na(Broth) & !is.na(Incubation_hr))
ps_all_broths <- prune_taxa(taxa_sums(ps_all_broths) > 0, ps_all_broths)

# Define function to create "target vs other" dataframe
target_vs_other_df <- function(ps, taxrank, target_values, target_label) {
  # Agglomerate at specified taxonomic rank
  ps_rank <- tax_glom(ps, taxrank = taxrank, NArm = FALSE)
  ps_rank <- prune_taxa(taxa_sums(ps_rank) > 0, ps_rank)
  
  # Transform to relative abundance
  ps_rel  <- transform_sample_counts(ps_rank, function(x) x / sum(x))

  # Convert to long format and categorize
  df <- psmelt(ps_rel) %>%
    mutate(
      Taxon = .data[[taxrank]],
      Taxon = ifelse(is.na(Taxon) | Taxon == "", "Unassigned", as.character(Taxon)),
      Focus = ifelse(Taxon %in% target_values, target_label, "Other")
    ) %>%
    group_by(Broth, Sample, Incubation_hr, Focus) %>%
    summarise(Abundance = sum(Abundance), .groups = "drop") %>%
    mutate(Focus = factor(Focus, levels = c(target_label, "Other")))

  df
}

# ===== Campylobacter Analysis =====
df_campylobacter <- target_vs_other_df(
  ps = ps_all_broths,
  taxrank = "Genus",
  target_values = c("Campylobacter"),
  target_label = "Campylobacter"
)

plot_campylobacter_enrichment <- ggplot(
  df_campylobacter,
  aes(x = Incubation_hr, y = Abundance, fill = Focus)
) +
  geom_col(width = 3, color = "white", linewidth = 0.25) +
  facet_wrap(~ Broth, ncol = 2) +
  scale_x_continuous(breaks = c(0, 4, 8, 12, 24)) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(
    values = c(
      "Campylobacter" = "#1b9e77",
      "Other" = "grey85"
    ),
    labels = c(
      expression(italic("Campylobacter")),
      "Other"
    )
  ) +
  guides(fill = guide_legend(ncol = 1)) +
  labs(
    x = "Incubation (h)",
    y = "Relative abundance",
    fill = NULL,
    title = expression(
      italic("Campylobacter") ~ "vs other bacteria (all enrichment broths)"
    )
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())

plot_campylobacter_enrichment

ggsave(
  "plot_Campylobacter_BB+CAT.png",
  plot  = plot_campylobacter_enrichment,
  width = 12,
  height = 8,
  dpi   = 600
)


library(phyloseq)
library(dplyr)
library(ggplot2)
library(scales)

# ---- Clean sample metadata ----
sample_metadata <- as.data.frame(sample_data(ps_enrichment_testing))

sample_metadata$Broth <- trimws(as.character(sample_metadata$Broth))
sample_metadata$Incubation_hr <- as.numeric(
  gsub("[^0-9.]", "", as.character(sample_metadata$Incubation))
)

sample_data(ps_enrichment_testing)$Broth <- sample_metadata$Broth
sample_data(ps_enrichment_testing)$Incubation_hr <- sample_metadata$Incubation_hr

# ---- Subset to valid samples ----
ps_pathogens <- subset_samples(
  ps_enrichment_testing,
  !is.na(Broth) & !is.na(Incubation_hr)
)
ps_pathogens <- prune_taxa(taxa_sums(ps_pathogens) > 0, ps_pathogens)

# ---- Agglomerate at Genus level ----
ps_genus <- tax_glom(ps_pathogens, taxrank = "Genus", NArm = FALSE)
ps_genus <- prune_taxa(taxa_sums(ps_genus) > 0, ps_genus)

# ---- Convert to relative abundance ----
ps_rel <- transform_sample_counts(ps_genus, function(x) x / sum(x))

# ---- Melt and classify taxa ----
df_pathogens <- psmelt(ps_rel) %>%
  mutate(
    Genus = trimws(as.character(Genus)),
    Genus = ifelse(is.na(Genus) | Genus == "", "Unassigned", Genus),
    Group = case_when(
      grepl("^Salmonella", Genus, ignore.case = TRUE)  ~ "Salmonella",
      grepl("^Escherichia", Genus, ignore.case = TRUE) ~ "Escherichia (E. coli)",
      TRUE                                              ~ "Other"
    )
  ) %>%
  group_by(Broth, Sample, Incubation_hr, Group) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop") %>%
  mutate(
    Group = factor(
      Group,
      levels = c("Salmonella", "Escherichia (E. coli)", "Other")
    )
  )

# ---- Plot ----
plot_pathogen_enrichment <- ggplot(
  df_pathogens,
  aes(x = Incubation_hr, y = Abundance, fill = Group)
) +
  geom_col(width = 0.6, color = "white", linewidth = 0.25) +
  facet_wrap(~ Broth, ncol = 2) +
  scale_x_continuous(breaks = c(0, 4, 8, 12, 24)) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(
    values = c(
      "Salmonella" = "#d95f02",
      "Escherichia (E. coli)" = "#1b9e77",
      "Other" = "grey85"
    ),
    labels = c(
      expression(italic("Salmonella")),
      expression(italic("Escherichia coli")),
      "Other"
    )
  ) +
  guides(fill = guide_legend(ncol = 1)) +
  labs(
    x = "Incubation (h)",
    y = "Relative abundance",
    fill = NULL,
    title = expression(
      italic("Salmonella") * ", " *
      italic("Escherichia coli") *
      ", and other bacteria across enrichment broths"
    )
  ) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    panel.border = element_blank()
  )

plot_pathogen_enrichment

# ---- Save ----
ggsave(
  "plot_pathogen_enrichment_all_broths.png",
  plot  = plot_pathogen_enrichment,
  width = 12,
  height = 8,
  dpi   = 600
)


   
```

```{r}

# Agglomerate at Genus level (keeps "Salmonella" if present)
ps_genus <- tax_glom(ps_enrichment_testing, taxrank = "Genus", NArm = FALSE)

# Melt to long format
dfg <- psmelt(ps_genus) %>%
  mutate(
    Broth = trimws(as.character(Broth)),
    Genus = ifelse(is.na(Genus) | Genus == "", "Unassigned", as.character(Genus))
  )

# Filter to BPW + RVS and Salmonella genus
df_sal <- dfg %>%
  filter(Genus == "Salmonella") %>%
  group_by(Broth) %>%
  summarise(
    n_samples_with_salmonella = sum(Abundance > 0, na.rm = TRUE),
    total_reads_salmonella = sum(Abundance, na.rm = TRUE),
    max_reads_in_a_sample = max(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

df_sal


```


```{r}
library(phyloseq)
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(scales)
library(grid)  # for unit()

# ---- 1) Clean metadata + subset to Bolton broths ----
md <- as.data.frame(sample_data(ps_enrichment_testing))
md$Broth <- trimws(as.character(md$Broth))
md$Incubation_hr <- as.numeric(gsub("[^0-9.]", "", as.character(md$Incubation)))
sample_data(ps_enrichment_testing)$Broth <- md$Broth
sample_data(ps_enrichment_testing)$Incubation_hr <- md$Incubation_hr

bolton_broths <- c("BF Bolton base", "BF Bolton base w CAT")
ps_bolton <- subset_samples(ps_enrichment_testing, Broth %in% bolton_broths)
ps_bolton <- prune_taxa(taxa_sums(ps_bolton) > 0, ps_bolton)

# ---- 2) Agglomerate to Species ----
ps_species <- tax_glom(ps_bolton, taxrank = "Species", NArm = FALSE)
ps_species <- prune_taxa(taxa_sums(ps_species) > 0, ps_species)

# ---- 3) Melt to long dataframe ----
df <- psmelt(ps_species) %>%
  mutate(
    Broth = trimws(as.character(Broth)),
    Incubation_hr = as.numeric(Incubation_hr),
    Species = ifelse(is.na(Species) | Species == "", "Unassigned", as.character(Species)),
    Genus = ifelse(is.na(Genus) | Genus == "", "Unassigned", as.character(Genus))
  )

# ---- 4) Compute relative abundance across all taxa ----
df <- df %>%
  group_by(Sample) %>%
  mutate(Abundance_rel = Abundance / sum(Abundance)) %>%
  ungroup()

# ---- 5) Keep top Campylobacter species; rest = Other ----
topN <- 8
top_species <- df %>%
  filter(Genus == "Campylobacter") %>%
  group_by(Species) %>%
  summarise(total = sum(Abundance_rel, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total)) %>%
  slice_head(n = topN) %>%
  pull(Species)

df <- df %>%
  mutate(
    Species2 = case_when(
      Genus == "Campylobacter" & Species %in% top_species ~ Species,
      TRUE ~ "Other"
    )
  )

# ---- 6) Order legend: top species first, Unassigned, then Other ----
species_order <- df %>%
  group_by(Species2) %>%
  summarise(total = sum(Abundance_rel, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total)) %>%
  pull(Species2)

# Make sure Unassigned and Other are last
species_order <- c(setdiff(species_order, c("Unassigned", "Other")), "Unassigned", "Other")
df$Species2 <- factor(df$Species2, levels = species_order)

# ---- 7) Color palette ----
n_species <- length(levels(df$Species2))
pal <- colorRampPalette(brewer.pal(12, "Paired"))(n_species)
names(pal) <- levels(df$Species2)
pal["Other"] <- "grey85"
pal["Unassigned"] <- "grey65"  # optional: distinguish from Other

# ---- 8) Plot stacked bars ----
plot_campy_species_bolton <- df %>%
  group_by(Broth, Sample, Incubation_hr, Species2) %>%
  summarise(Abundance_rel = sum(Abundance_rel), .groups = "drop") %>%
  ggplot(aes(x = Incubation_hr, y = Abundance_rel, fill = Species2)) +
  geom_col(width = 0.9, color = "white", linewidth = 0.25) +
  facet_wrap(~ Broth, ncol = 1) +
  scale_x_continuous(breaks = c(0, 4, 8, 12, 24)) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(
    values = pal,
    drop = FALSE,
    labels = function(x) {
      ifelse(
        x %in% c("Other", "Unassigned"),
        x,
        parse(text = paste0("italic('", x, "')"))
      )
    }
  ) +
  labs(
    x = "Incubation (h)",
    y = "Relative abundance (all bacteria)",
    fill = "Species",
    title = "Relative abundance of" ~ 
      italic("Campylobacter") ~ "species in Bolton broths"
  ) +
  theme_bw(base_size = 16) +
  theme(
    panel.grid = element_blank(),
    legend.key.height = unit(0.45, "cm")
  )


plot_campy_species_bolton

ggsave(
  "plot_Campylobacter_BB_CAT_RA_(all-bacteria).png",
  plot  = plot_campy_species_bolton,
  width = 12,
  height = 8,
  dpi   = 600
)

```

```{r Enrichment: Campylobacter species vs all other bacteria (Bolton broths) + results table}

# ---------- 1) Clean metadata + subset to Bolton broths ----------
md <- as.data.frame(sample_data(ps_enrichment_testing))
md$Broth <- trimws(as.character(md$Broth))
md$Incubation_hr <- as.numeric(gsub("[^0-9.]", "", as.character(md$Incubation)))
sample_data(ps_enrichment_testing)$Broth <- md$Broth
sample_data(ps_enrichment_testing)$Incubation_hr <- md$Incubation_hr

bolton_broths <- c("BF Bolton base", "BF Bolton base w CAT")

ps_bolton <- subset_samples(ps_enrichment_testing, Broth %in% bolton_broths)
ps_bolton <- prune_taxa(taxa_sums(ps_bolton) > 0, ps_bolton)

# ---------- 2) Make TWO views: raw counts (for table) + relative abundance (for plot) ----------
# Agglomerate to Species for both
ps_sp_counts <- tax_glom(ps_bolton, taxrank = "Species", NArm = FALSE)
ps_sp_counts <- prune_taxa(taxa_sums(ps_sp_counts) > 0, ps_sp_counts)

ps_sp_rel <- transform_sample_counts(ps_sp_counts, function(x) x / sum(x))  # relative abundance of ALL bacteria

df_counts <- psmelt(ps_sp_counts) %>%
  mutate(
    Broth = trimws(as.character(Broth)),
    Incubation_hr = as.numeric(Incubation_hr),
    Genus = ifelse(is.na(Genus) | Genus == "", "Unassigned", as.character(Genus)),
    Species = ifelse(is.na(Species) | Species == "", "Unassigned", as.character(Species))
  )

df_rel <- psmelt(ps_sp_rel) %>%
  mutate(
    Broth = trimws(as.character(Broth)),
    Incubation_hr = as.numeric(Incubation_hr),
    Genus = ifelse(is.na(Genus) | Genus == "", "Unassigned", as.character(Genus)),
    Species = ifelse(is.na(Species) | Species == "", "Unassigned", as.character(Species))
  )

# ---------- 3) Determine "most common" Campylobacter species (by TOTAL READS, not RA) ----------
# This prevents rare species from dominating because of normalization quirks
topN <- 8  # change as desired

top_campy_species <- df_counts %>%
  filter(Genus == "Campylobacter") %>%
  group_by(Species) %>%
  summarise(total_reads = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_reads)) %>%
  slice_head(n = topN) %>%
  pull(Species)

# ---------- 4) Build plot dataframe: top Campylobacter species, everything else = Other ----------
df_plot <- df_rel %>%
  mutate(
    Focus = case_when(
      Genus == "Campylobacter" & Species %in% top_campy_species ~ Species,
      TRUE ~ "Other"
    )
  ) %>%
  group_by(Broth, Sample, Incubation_hr, Focus) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop")

# Order legend: top species first, Other last (ordered by total abundance in plot space)
focus_order <- df_plot %>%
  group_by(Focus) %>%
  summarise(tot = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(tot)) %>%
  pull(Focus)
focus_order <- c(setdiff(focus_order, "Other"), "Other")
df_plot$Focus <- factor(df_plot$Focus, levels = focus_order)

# Colors: species get distinct colors; Other grey
n_focus <- length(levels(df_plot$Focus))
pal <- colorRampPalette(brewer.pal(12, "Paired"))(n_focus)
names(pal) <- levels(df_plot$Focus)
pal["Other"] <- "grey85"

plot_campy_species_allbacteria <- ggplot(df_plot, aes(x = Incubation_hr, y = Abundance, fill = Focus)) +
  geom_col(width = 2, color = "white", linewidth = 0.25) +
  facet_wrap(~Broth, ncol = 1) +
  scale_x_continuous(breaks = c(0, 4, 8, 12, 24)) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = pal, drop = FALSE) +
  labs(
    x = "Incubation (h)",
    y = "Relative abundance (all bacteria)",
    fill = "Campylobacter species",
    title = paste0("Campylobacter species (top ", topN, ") vs all other bacteria (Bolton broths)")
  ) +
  theme_bw(base_size = 16) +
  theme(
    panel.grid = element_blank(),
    legend.key.height = unit(0.45, "cm")
  )

plot_campy_species_allbacteria

# Optional save
ggsave(
  filename = "16S_Campylobacter_species_vs_other_allbacteria_Bolton.png",
  plot = plot_campy_species_allbacteria,
  width = 10, height = 8, dpi = 300
)

# ---------- 5) Results table: species, broth, incubation time, read count ----------
# (Shows even low-count species; can filter later if too big)

campy_table <- df_counts %>%
  filter(Genus == "Campylobacter") %>%
  group_by(Broth, Incubation_hr, Species) %>%
  summarise(Read_Count = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(Broth, Incubation_hr, desc(Read_Count))

campy_table

# Optional: write to CSV
write.csv(campy_table, "Campylobacter_species_read_counts_Bolton_broths.csv", row.names = FALSE)

# Optional: a wide version for easier viewing (Species x time within each broth)
campy_table_wide <- campy_table %>%
  tidyr::pivot_wider(names_from = Incubation_hr, values_from = Read_Count, values_fill = 0) %>%
  arrange(Broth, desc(rowSums(across(where(is.numeric)))))

campy_table_wide
```


##Stats - Bolton broths - Campylobacter

```{r}

library(dplyr)
library(tidyr)
library(stringr)

# ---- 1) Summarize total Campylobacter relative abundance per sample ----
campy_summary <- df %>%
  group_by(Broth, Sample, Incubation_hr) %>%
  summarise(Campylobacter_rel = sum(Abundance_rel[Genus == "Campylobacter"], na.rm = TRUE),
            .groups = "drop")

# ---- 2) Create a common replicate ID ----
campy_summary <- campy_summary %>%
  mutate(
    Replicate = str_remove(Sample, "\\d+$")  # remove trailing digits (the time point)
  )
campy_summary

# ---- 3) Filter only 0h and 24h ----
campy_0_24 <- campy_summary %>%
  filter(Incubation_hr %in% c(0, 24))

campy_0_24

# ---- 4) Reshape to wide format using replicate ID ----
campy_wide <- campy_0_24 %>%
  select(Broth, Replicate, Incubation_hr, Campylobacter_rel) %>%
  pivot_wider(names_from = Incubation_hr,
              values_from = Campylobacter_rel,
              names_prefix = "h") %>%
  drop_na(h0, h24)  # only keep replicates with both 0h and 24h

campy_wide

# ---- 5) Check number of paired replicates per broth ----
campy_wide %>% group_by(Broth) %>% summarise(n_pairs = n())


ggplot(campy_0_24, aes(x = Incubation_hr, y = Campylobacter_rel, group = Replicate)) +
  geom_line(aes(color = Replicate), size = 1) +
  geom_point(size = 3) +
  scale_x_continuous(breaks = c(0, 24)) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    x = "Incubation (h)",
    y = "Relative abundance (all bacteria)",
    title = "Campylobacter enrichment in Bolton broths"
  ) +
  theme_bw(base_size = 14) +
  theme(legend.position = "none")


```




#Host-DNA Depletion Analysis

##Host-DNA depletion samples

```{r}
#Filtering samples by condition - Host-DNA depletion
samples.out

Hostdepletion_originalnames <- c("HCh1", "HCh1ND", "HCh2", "HCh2ND", "HCh3", "HCh3ND", "HChCAT1", "HChCAT1ND", "HChCAT2", "HChCAT2ND", "HChCAT3", "HChCAT3ND", "HChRVS1", "HChRVS1ND", "HChRVS2", "HChRVS2ND", "HChRVS3", "HChRVS3ND", "NHCh1", "NHCh2", "NHCh3", "NHChCAT1", "NHChCAT2", "NHChCAT3", "NHChRVS1", "NHChRVS2", "NHChRVS3", "NHChRVSB")

Hostdepletion_originalnames

hostdepletion <- Hostdepletion_originalnames[Hostdepletion_originalnames %in% samples.out]

hostdepletion

hostdepletion <- gsub("^NH", "No Host-DNA depletion ", hostdepletion)
hostdepletion <- gsub("^H", "Host-DNA depletion ", hostdepletion)

hostdepletion <- gsub("Ch", "Cheesecloth ", hostdepletion)

hostdepletion <- gsub("F", "Filtered ", hostdepletion, fixed = TRUE)

hostdepletion <- gsub("B", "Blank", hostdepletion, fixed = TRUE)

hostdepletion <- gsub("CAT", "BFBB w CAT ", hostdepletion)

hostdepletion <- gsub("Ce", "Centrifuge ", hostdepletion)

hostdepletion <- gsub("RVS", "RVS ", hostdepletion)

hostdepletion <- gsub("ND", "_No-dilution", hostdepletion)

hostdepletion
```

#Create data table for Host-DNA depletion

```{r}
# Extract Host-DNA depletion status
host_depletion_status <- str_extract(hostdepletion, "^(No Host-DNA depletion|Host-DNA depletion)")

host_depletion_status

  # Extracting 'No-dilution' status
  no_dilution <- ifelse(grepl("_No-dilution", hostdepletion), "Diluted", "Not Diluted")
no_dilution

filtering_method <- str_extract(hostdepletion, "Cheesecloth|Filtered Centrifuge|Filtered")
filtering_method

incubation <- str_extract(hostdepletion, "BFBB w CAT|RVS")
incubation

# Replace NA with "No Incubation"
incubation <- ifelse(is.na(incubation), "No Incubation", incubation)

Blank_No <- str_extract(hostdepletion, "1|2|3|Blank")

samdf2 <- data.frame("Name"=Hostdepletion_originalnames, "Sample_Name"=hostdepletion, "Host-DNA Depletion"=host_depletion_status, "Method of Filtering"=filtering_method, Incubation=incubation, Dilution=no_dilution, "Sample Number/Type"=Blank_No)

samdf2

rownames(samdf2) <- (Hostdepletion_originalnames)

ps_hostDNAdep <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows=FALSE), sample_data(samdf2), tax_table(taxa))

### plotting diversity by day using the Shannon and Simpson measures

plot_richness(ps_hostDNAdep, x="Host.DNA.Depletion", measures=c("Shannon", "Simpson"), color="Method.of.Filtering")

```

#Host-DNA depletion testing - filtered and filtered centrifuge samples

```{r}

# Select top 20 most abundant taxa
ps_hostDNAdep <- tax_glom(ps_hostDNAdep, taxrank = "Family")

top20_hostDNAdep <- names(sort(taxa_sums(ps_hostDNAdep), decreasing=TRUE))[1:20]

# Normalize abundances
ps_hostDNAdep.top20 <- transform_sample_counts(ps_hostDNAdep, function(OTU) OTU/sum(OTU))

# Keep only top 20 taxa
ps_hostDNAdep.top20 <- prune_taxa(top20_hostDNAdep, ps_hostDNAdep.top20)

# Aggregate at Family level to combine taxa from the same Family
ps_hostDNAdep.top20_family <- tax_glom(ps_hostDNAdep.top20, taxrank = "Family")


ps_no_cheesecloth <- subset_samples(ps_hostDNAdep.top20_family, 
                                    Method.of.Filtering != "Cheesecloth")



#Combine labels in sample number and dilution column 
sample_data(ps_no_cheesecloth)$Sample_both <- paste(
  sample_data(ps_no_cheesecloth)$Sample.Number.Type, 
  sample_data(ps_no_cheesecloth)$Dilution, 
  sep = " - "
)
ps_no_cheesecloth


# Generate the bar plot
p4 <- plot_bar(ps_no_cheesecloth, 
              x = "Sample_both", fill = "Family") + 
     facet_wrap(~ Host.DNA.Depletion + Method.of.Filtering, scales = "free_x") + 
     theme(axis.text.x = element_text(angle = 90, hjust = 0.5))
p4


jpeg(file="taxa20_hostDNAdepletion-wocheesecloth.jpg", height=1000, width=1800)
# Final plot with clearer color grouping
p4 +  # outline between segments
  theme_bw(base_size = 16) +  # Increase base text size for everything
  theme(
    axis.text.x = element_text(size = 14, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 16, face = "bold"),
    axis.title.y = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 16),  # for facet labels
    panel.grid = element_blank()
  )
dev.off()

ps_no_cheesecloth_ra <- transform_sample_counts(ps_no_cheesecloth, function(x) 100 * x / sum(x))

p_ra_nocheesecloth <- plot_bar(ps_no_cheesecloth_ra, 
              x = "Sample_both", fill = "Family") + 
     facet_wrap(~ Host.DNA.Depletion + Method.of.Filtering, scales = "free_x") + 
     theme(axis.text.x = element_text(angle = 90, hjust = 0.5))
p_ra_nocheesecloth

#Relative abundance plot
jpeg(file="taxa20_hostDNAdepletion-nocheesecloth_ra.jpg", height=1000, width=1800)
p_ra_nocheesecloth +  # outline between segments
  theme_bw(base_size = 16) +  # Increase base text size for everything
  theme(
    axis.text.x = element_text(size = 14, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 16, face = "bold"),
    axis.title.y = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 16),  # for facet labels
    panel.grid = element_blank()
  )
dev.off()

```

#Host-DNA depletion samples - cheesecloth only RA

```{r}
ps_hostDNAdep.top20_family <- tax_glom(ps_hostDNAdep.top20, taxrank = "Family")

ps_cheesecloth <- subset_samples(ps_hostDNAdep.top20_family, 
                                    Method.of.Filtering != "Filtered")

ps_cheesecloth <- subset_samples(ps_cheesecloth, 
                                    Method.of.Filtering != "Filtered Centrifuge")

sample_data(ps_cheesecloth)$Sample_both <- paste(
  sample_data(ps_cheesecloth)$Sample.Number.Type, 
  sample_data(ps_cheesecloth)$Dilution, 
  sep = " - "
)

p_cheesecloth <- plot_bar(ps_cheesecloth, 
              x = "Sample_both", fill = "Family") + 
     facet_wrap(~ Host.DNA.Depletion + Method.of.Filtering, scales = "free_x") + 
     theme(axis.text.x = element_text(angle = 90, hjust = 0.5))
p_cheesecloth


jpeg(file="taxa20_hostDNAdepletion-cheesecloth.jpg", height=1000, width=1800)
p_cheesecloth +  # outline between segments
  theme_bw(base_size = 16) +  # Increase base text size for everything
  theme(
    axis.text.x = element_text(size = 14, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 16, face = "bold"),
    axis.title.y = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 16),  # for facet labels
    panel.grid = element_blank()
  )
dev.off()


ps_cheesecloth_ra <- transform_sample_counts(ps_cheesecloth, function(x) 100 * x / sum(x))


sample_data(ps_cheesecloth_ra)$Sample_both <- paste(
  sample_data(ps_cheesecloth_ra)$Sample.Number.Type, 
  sample_data(ps_cheesecloth_ra)$Dilution, 
  sep = " - "
)
ps_cheesecloth_ra


# Generate the bar plot for cheesecloth relative abundance
p_cheesecloth_ra <- plot_bar(ps_cheesecloth_ra, 
              x = "Sample_both", fill = "Family") + 
     facet_wrap(~ Host.DNA.Depletion + Method.of.Filtering, scales = "free_x") + 
     theme(axis.text.x = element_text(angle = 90, hjust = 0.5))
p_cheesecloth_ra


jpeg(file="taxa20_hostDNAdepletion-justcheesecloth_ra.jpg", height=1000, width=1800)
# Final plot with clearer color grouping
p_cheesecloth_ra +  # outline between segments
  theme_bw(base_size = 16) +  
  geom_bar(stat = "identity", position = "stack", color = "black") + 
  theme(
    axis.text.x = element_text(size = 14, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 16, face = "bold"),
    axis.title.y = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 16),  # for facet labels
    panel.grid = element_blank()
  )
dev.off()

```

#Host-DNA depletion samples - enriched -

```{r}
ps_cheesecloth_incubation <- subset_samples(ps_hostDNAdep.top20_family, 
                                    Method.of.Filtering != "Filtered")

ps_cheesecloth_incubation <- subset_samples(ps_cheesecloth_incubation, 
                                    Method.of.Filtering != "Filtered Centrifuge")

ps_cheesecloth_incubation <- subset_samples(ps_cheesecloth_incubation, 
                                    Incubation != "No Incubation")


sample_data(ps_cheesecloth_incubation)$Sample_both <- paste(
  sample_data(ps_cheesecloth_incubation)$Sample.Number.Type,
  sample_data(ps_cheesecloth_incubation)$Dilution, 
  sep = " - "
)
ps_cheesecloth_incubation


ps_cheesecloth_incubation <- tax_glom(ps_cheesecloth_incubation, taxrank = "Family")

# Generate the bar plot
p_cheesecloth_incubation <- plot_bar(ps_cheesecloth_incubation, 
              x = "Sample_both", fill = "Family") + 
     facet_wrap(~ Host.DNA.Depletion + Incubation, scales = "free_x") + 
     theme(axis.text.x = element_text(angle = 90, hjust = 0.5))
p_cheesecloth_incubation

jpeg(file="taxa20_hostDNAdepletion-enriched.jpg", height=1000, width=1800)
# Final plot with clearer color grouping
p_cheesecloth_incubation +  # outline between segments
  theme_bw(base_size = 16) +  
  geom_bar(stat = "identity", position = "stack", color = "black") + 
  theme(
    axis.text.x = element_text(size = 14, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 16, face = "bold"),
    axis.title.y = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 16),  # for facet labels
    panel.grid = element_blank()
  )
dev.off()

ps_cheesecloth_incubation_ra <- transform_sample_counts(ps_cheesecloth_incubation, function(x) 100 * x / sum(x))

# Generate the bar plot
p_cheesecloth_incubation_ra <- plot_bar(ps_cheesecloth_incubation_ra, 
              x = "Sample_both", fill = "Family") + 
     facet_wrap(~ Host.DNA.Depletion + Incubation, scales = "free_x") + 
     theme(axis.text.x = element_text(angle = 90, hjust = 0.5))
p_cheesecloth_incubation_ra

jpeg(file="taxa20_hostDNAdepletion-enriched_ra.jpg", height=1000, width=1800)
# Final plot with clearer color grouping
p_cheesecloth_incubation_ra +  # outline between segments
  theme_bw(base_size = 16) +  
  geom_bar(stat = "identity", position = "stack", color = "black") + 
  theme(
    axis.text.x = element_text(size = 14, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 16, face = "bold"),
    axis.title.y = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 16),  # for facet labels
    panel.grid = element_blank()
  )
dev.off()


#Check for Salmonella  in RVS 
subset_salmonella_host <- subset_taxa(ps_cheesecloth_incubation, Phylum %in% c("Pseudomonadota"))
as.data.frame(tax_table(subset_salmonella_host))

```

#DNA Extraction kit comparisons

```{r}
#Filtering samples by condition - Host-DNA depletion
samples.out

dnakits_originalnames <- c("PRMgDNAUn", "PrMigDNA1", "PrMigDNA2", "PrMigDNA3", "QDnBT1", "QDnBT2", "QDnBT3", "QDnBTSp", "QDnBTUn", "ZymoMiPr1", "ZymoMiPr2", "ZymoMiPr3", "ZymoMiPrSp", "ZymoMIPrUn")

dnakits_originalnames

DNAkits <- dnakits_originalnames[dnakits_originalnames %in% samples.out]

DNAkits

DNAkits <- gsub("^PrMigDNA", "Presto Mini gDNA ", DNAkits)
DNAkits <- gsub("^PRMgDNA", "Presto Mini gDNA ", DNAkits)

DNAkits <- gsub("^QDnBT", "Qiagen DNeasy Blood & Tissue ", DNAkits)
DNAkits <- gsub("^ZymoMiPr", "ZymoResearch DNA Miniprep ", DNAkits)
DNAkits <- gsub("^ZymoMIPr", "ZymoResearch DNA Miniprep ", DNAkits)

DNAkits <- gsub("Sp", "Spiked", DNAkits)
DNAkits <- gsub("Un", "Unspiked", DNAkits)

DNAkits

```

#Create data table for Host-DNA depletion

```{r}
DNAkit_type <- str_extract(DNAkits, "^(ZymoResearch DNA Miniprep|Qiagen DNeasy Blood & Tissue|Presto Mini gDNA)")

DNAkit_type

Spiking <- str_extract(DNAkits, "Unspiked|Spiked")
Spiking
Spiking <- ifelse(is.na(Spiking), "Chicken Rinse", Spiking)

Number <- str_extract(DNAkits, "1|2|3")
Number
Number <- ifelse(is.na(Number), "Control", Number)

samdf3 <- data.frame("DNA Kit"=DNAkit_type, "Sample"=Spiking, "Number/Type"=Number)

samdf3

rownames(samdf3) <- (dnakits_originalnames)


DNAkits2 <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows=FALSE), sample_data(samdf3), tax_table(taxa))

DNAkits2

sample_data(DNAkits2)$Name <- paste(
  sample_data(DNAkits2)$Number,
  sample_data(DNAkits2)$Sample, 
  sep = " - "
)

```

#Host-DNA depletion testing - filtered and filtered centrifuge samples

```{r}
DNAkits2 <- tax_glom(DNAkits2, taxrank = "Family")

# Select top 20 most abundant taxa
top20_dnakits <- names(sort(taxa_sums(DNAkits2), decreasing=TRUE))[1:20]

# Normalize abundances
DNAkits2.top20 <- transform_sample_counts(DNAkits2, function(OTU) OTU/sum(OTU))
DNAkits2.top20

# Keep only top 20 taxa
DNAkits2.top20 <- prune_taxa(top20_dnakits, DNAkits2.top20)

# Aggregate at Family level to combine taxa from the same Family
DNAkits2.top20_family <- tax_glom(DNAkits2.top20, taxrank = "Family")

DNAkits2.top20_family

colours <- c("#72e5ef", "#c23165", "#aef815", "#6f3a93", "#56a221", "#cf60f3", "#2cf52b", "#fe16f4", "#93f993", "#bf7d9a", "#1c5f1e", "#f2c8e8", "#145a6a", "#d5ec9e", "#7c08c5", "#f3d426", "#0494fb", "#96a467", "#8e80fb", "#0ba47e")

colours

# Generate the bar plot
p_DNA <- plot_bar(DNAkits2.top20_family, 
              x = "Name", fill = "Family") + 
  facet_wrap(~DNA.Kit, scales = "free_x") +
     theme(axis.text.x = element_text(angle = 90, hjust = 0.5))

p_DNA

jpeg(file="taxa20_DNAkits.jpg", height=1000, width=1800)
# Final plot with clearer color grouping
p_DNA +  # outline between segments
  theme_bw(base_size = 16) +  
  geom_bar(stat = "identity", position = "stack", color = "black") + 
  theme(
    axis.text.x = element_text(size = 14, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 16, face = "bold"),
    axis.title.y = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 16),  # for facet labels
    panel.grid = element_blank()
  ) +
  scale_fill_manual(values=colours)
dev.off()

```

```{r}

ps_DNAkits_ra <- transform_sample_counts(DNAkits2.top20_family, function(x) 100 * x / sum(x))

p_DNAkits_ra <- plot_bar(ps_DNAkits_ra, 
              x = "Name", fill = "Family") + 
  facet_wrap(~DNA.Kit, scales = "free_x") +
     theme(axis.text.x = element_text(angle = 90, hjust = 0.5))

p_DNAkits_ra


#Relative abundance plot
jpeg(file="taxa20_DNAkits_ra.jpg", height=1000, width=1800)
p_DNAkits_ra +  # outline between segments
  theme_bw(base_size = 16) +  # Increase base text size for everything
  theme(
    axis.text.x = element_text(size = 14, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 16, face = "bold"),
    axis.title.y = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 16),  # for facet labels
    panel.grid = element_blank()
  )
dev.off()

```

```{r}
head(track)

track_data <- data.frame(track)
track_data


library(pheatmap)


track_data


pheatmap(track_data,
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         color = colorRampPalette(c("white", "blue"))(100),
         main = "Read Counts at Each Step",
         fontsize_row = 10,     # adjust text size for row labels
         fontsize_col = 10)     # adjust text size for column labels



# Convert to a matrix (1 row or 1 column)
nonchim_matrix <- as.matrix(track_data$nonchim)

# Optionally set row/column names
rownames(nonchim_matrix) <- rownames(track_data)
colnames(nonchim_matrix) <- "nonchim"

# Check rownames and colnames
rownames(nonchim_matrix)
colnames(nonchim_matrix)

# If your sample names are the rownames of track_data
nonchim_matrix <- as.matrix(track_data$nonchim)

# Add rownames if missing
rownames(nonchim_matrix) <- rownames(track_data)

# Or, if you have sample names in another column (say, "SampleID")
rownames(nonchim_matrix) <- track_data$SampleID



# Then plot!
pheatmap(t(nonchim_matrix), 
         cluster_rows = FALSE, 
         cluster_cols = FALSE,
         color = colorRampPalette(c("white", "#23238c"))(100),
         labels_row = colnames(nonchim_matrix),
         labels_col = rownames(nonchim_matrix))



```

```{r}

library(aplot)

combined_plot <- pheatmap %>% 
  insert_right(p, width = 0.25)

```

```{r}
library(ggplot2)
library(reshape2)
library(patchwork)  # for combining ggplots

# Assuming nonchim_matrix is already defined
# Let's transpose it, since pheatmap(t(nonchim_matrix)) was used
heatmap_data <- t(nonchim_matrix)

# Convert matrix to long format for ggplot2
df_heatmap <- melt(heatmap_data)

# Rename for clarity
colnames(df_heatmap) <- c("Row", "Col", "Value")

# Create the heatmap with ggplot2
p_heatmap <- ggplot(df_heatmap, aes(x = Col, y = Row, fill = Value)) +
  geom_tile() +
  scale_fill_gradientn(colors = c("white", "#23238c")) +
  theme_minimal() +
  labs(x = NULL, y = NULL) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    axis.title = element_blank()
  )

p_heatmap


```

```{r}
library(phyloseq)
library(ggplot2)
library(dplyr)
library(tidyr)

# Convert phyloseq object to data.frame
df <- phyloseq::psmelt(ps_enrichment_testing.top20_enrichment_testing)

# Inspect the structure
head(df)

gg_p <- ggplot(df, aes(x = Incubation, y = Abundance, fill = Family)) +
  geom_bar(stat = "identity", position = "stack", color = "black", linewidth=0.3) +
  facet_wrap(~Broth, scales = "free_x") +  # Facet by Broth
  theme_bw(base_size = 16) +  # Base font size
  theme(
    axis.text.x = element_text(size = 14, angle = 0, hjust = 0.5),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 16, face = "bold"),
    axis.title.y = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 16),
    panel.grid = element_blank()
  ) +
  labs(
    x = "Incubation",
    y = "Abundance",
    fill = "Family"
  )
gg_p$data$Incubation <- factor(gg_p$data$Incubation, levels = sort(as.numeric(unique(gg_p$data$Incubation))))

gg_p


```

```{r}
# Load required libraries
library(phyloseq)
library(ggplot2)
library(reshape2)

# Extract OTU table (samples as columns)
otu_mat <- as(otu_table(ps_enrichment_testing.top20_enrichment_testing), "matrix")

# Make sure taxa are rows
if(!taxa_are_rows(ps_enrichment_testing.top20_enrichment_testing)) {
  otu_mat <- t(otu_mat)
}

# Inspect it
dim(otu_mat)
head(otu_mat)

```

```{r}
# Extract sample metadata from the phyloseq object
sample_metadata <- as.data.frame(sample_data(ps_enrichment_testing.top20_enrichment_testing))
head(sample_metadata)

# Add the Sample column to sample_metadata to make joining easier
sample_metadata$Sample <- rownames(sample_metadata)

# Join df_long with sample metadata on the Sample column
df_long <- df_long %>%
  left_join(sample_metadata, by = "Sample")

head(df_long)

# Summarize total reads by Sample, ignoring Family
df_total_reads <- df_long %>%
  group_by(Broth, Incubation) %>%
  summarise(Total_Reads = sum(Reads, na.rm = TRUE))

# View the summarized data
head(df_total_reads)


df_long$Incubation <- factor(df_long$Incubation, levels = sort(unique(df_long$Incubation)))

```

```{r}

df_long$Reads_log <- log10(df_long$Reads + 1)

# Extract the prefix (letters) and numeric parts
df_long$Prefix <- gsub("[0-9]", "", df_long$Sample)
df_long$Sample_num <- as.numeric(gsub("[A-Za-z]+", "", df_long$Sample))

# Arrange the data by Prefix and then Sample_num
df_long <- df_long %>%
  arrange(Prefix, Sample_num)

# Now reorder the Sample factor based on the sorted order
df_long$Sample <- factor(df_long$Sample, levels = unique(df_long$Sample))


# Now plot the heatmap with ordered sample names
p_heatmap <- ggplot(df_long, aes(x = Sample, y = Family, fill = Reads)) +
  geom_tile(color = "white") +  # White gridlines between tiles
  scale_fill_gradient(low = "white", high = "#23238c") +  # Customize colors
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    axis.text.y = element_text(size = 12),
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12)
  ) +
  labs(x = "Sample", y = "Family", fill = "Read Count")

# Show the heatmap
p_heatmap

```

```{r}

combined_plot <- p_heatmap + gg_p + plot_layout(ncol = 2)

combined_plot

```

```{r}
library(dplyr)

# Summarize total reads per sample
sample_reads <- df_long %>%
  group_by(Sample) %>%
  summarize(Total_Reads = sum(Reads)) %>%
  arrange(Total_Reads)

# View the summarized table
sample_reads

sample_reads$Sample <- factor(sample_reads$Sample, levels = sample_reads$Sample)

library(ggplot2)

p_total_reads_heatmap <- ggplot(sample_reads, aes(x = Sample, y = 1, fill = Total_Reads)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "#23238c") +  # Adjust colors if needed
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    axis.text.y = element_blank(),  # No y-axis because it's just one row
    axis.title.y = element_blank(),
    panel.grid = element_blank()
  ) +
  labs(x = "Sample", fill = "Total Reads", title = "Total Read Count per Sample")

p_total_reads_heatmap

```

```{r}
library(patchwork)

# Assume you already have:
# p_main = your main ggplot (e.g., bar plot, heatmap of families)
# p_total_reads_heatmap = the heatmap or barplot of total reads per sample

# Combine them vertically
combined_plot_totalread <- gg_p / p_total_reads_heatmap  # '/' stacks them vertically

# Show the plot
combined_plot_totalread




```

```{r}
library(ggplot2)
library(patchwork)

# Example: Clean top plot (remove x-axis)
p_main_clean <- gg_p +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )

# Example: Clean heatmap (remove y axis, optional)
p_heatmap_clean <- p_total_reads_heatmap +
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

# Combine with patchwork, sharing the x-axis
combined_plot <- p_main_clean / p_heatmap_clean +
  plot_layout(heights = c(4, 1))  # Control how much space each plot gets

combined_plot

library(cowplot)

aligned_plots <- align_plots(p_main_clean, p_heatmap_clean, align = 'v', axis = 'lr')

combined_plot <- plot_grid(
  aligned_plots[[1]],
  aligned_plots[[2]],
  ncol = 1,
  rel_heights = c(4, 1)
)

combined_plot

```
